---
title: 'Reproducible Research Assessment 2: Impact of severe weather events on US
  population health and economy'
author: "Rosalind"
date: "21 December, 2014"
output: pdf_document
---

##Synopsis

Aim: The overarching aim of this report is to understand the impact of severe weather events on the US population health and economy. 

Data source: Findings from this report is based on the NOAA Storm Database (period: 1950 to 2011) 

## Settings
```{r}
#Make code visible, set working directory and read library
echo = TRUE
setwd("~/RepData_PeerAssessment2")
library(ggplot2)
library(plyr)
library(utils)
require(gridExtra)
```

## Data exploration
```{r}
data <- read.csv(file="data/StormData.csv", header=T, sep=",", quote="", as.is=TRUE)
summary(data)
```

```{r}
if (!"StormData" %in% ls()) {
    Data <- read.csv("data/StormData.csv", sep = ",")
}
dim(Data)
head(Data, n = 2)
```
The total number of rows and columns are 902297 and 37 respectively.

The events started from 1950 and end in 2011. There were few recorded events in the earlier years due to the lack of good records. While the records were more complete in the recent years. 

```{r}
if (dim(Data)[2] == 37) {
    Data$year <- as.numeric(format(as.Date(Data$BGN_DATE, format = "%m/%d/%Y %H:%M:%S"), "%Y"))
}
hist(Data$year, breaks = 30)
```  

Based on the graph, the number of events tracked start to increase significantly from 1995. Hence, 1990 to 2011 will be use to make most of the data. 

```{r}
storm <- Data[Data$year >= 1995, ]
dim(storm)
```
With this, the total number of rows and column 681500 and 38 respectively.

#### Impact on Population Health
To identify the no. of **fatalities** and **injuries** that are caused by the severe weather events, we list the top 10 most severe types of weather events.

```{r}
sortHelper <- function(fieldName, top = 10, dataset = Data) {
    index <- which(colnames(dataset) == fieldName)
    field <- aggregate(dataset[, index], by = list(dataset$EVTYPE), FUN = "sum")
    names(field) <- c("EVTYPE", fieldName)
    field <- arrange(field, field[, 2], decreasing = T)
    field <- head(field, n = top)
    field <- within(field, EVTYPE <- factor(x = EVTYPE, levels = field$EVTYPE))
    return(field)
}

fatalities <- sortHelper("FATALITIES", dataset = storm)
injuries <- sortHelper("INJURIES", dataset = storm)
```

#### Impact on Economy
We will convert the **property damage** and **crop damage** data into comparable numerical forms. Both `PROPDMGEXP` and `CROPDMGEXP` columns record a multiplier for each observation where we have Hundred (H), Thousand (K), Million (M) and Billion (B).

```{r}
convertHelper <- function(dataset = storm, fieldName, newFieldName) {
    totalLen <- dim(dataset)[2]
    index <- which(colnames(dataset) == fieldName)
    dataset[, index] <- as.character(dataset[, index])
    logic <- !is.na(toupper(dataset[, index]))
    dataset[logic & toupper(dataset[, index]) == "B", index] <- "9"
    dataset[logic & toupper(dataset[, index]) == "M", index] <- "6"
    dataset[logic & toupper(dataset[, index]) == "K", index] <- "3"
    dataset[logic & toupper(dataset[, index]) == "H", index] <- "2"
    dataset[logic & toupper(dataset[, index]) == "", index] <- "0"
    dataset[, index] <- as.numeric(dataset[, index])
    dataset[is.na(dataset[, index]), index] <- 0
    dataset <- cbind(dataset, dataset[, index - 1] * 10^dataset[, index])
    names(dataset)[totalLen + 1] <- newFieldName
    return(dataset)
}

storm <- convertHelper(storm, "PROPDMGEXP", "propertyDamage")
storm <- convertHelper(storm, "CROPDMGEXP", "cropDamage")
names(storm)
options(scipen=999)
property <- sortHelper("propertyDamage", dataset = storm)
crop <- sortHelper("cropDamage", dataset = storm)
```

### Findings
As for the impact on public health, we have the lists of severe weather events by the no. of people badly injured.

```{r}
fatalities
injuries
```
Below illustrates the total fatalities and injuries caused by the severe weather events. 
```{r}
fatalitiesPlot <- qplot(EVTYPE, data = fatalities, weight = FATALITIES, geom = "bar", binwidth = 1) + 
    scale_y_continuous("No of Fatalities") + 
    theme(axis.text.x = element_text(angle = 45, 
    hjust = 1)) + xlab("Severe Weather Type") + 
    ggtitle("Total Fatalities by Severe Weather\n Events in the U.S.\n from 1995 - 2011")
injuriesPlot <- qplot(EVTYPE, data = injuries, weight = INJURIES, geom = "bar", binwidth = 1) + 
    scale_y_continuous("Number of Injuries") + 
    theme(axis.text.x = element_text(angle = 45, 
    hjust = 1)) + xlab("Severe Weather Type") + 
    ggtitle("Total Injuries by Severe Weather\n Events in the U.S.\n from 1995 - 2011")
grid.arrange(fatalitiesPlot, injuriesPlot, ncol = 2)
```  

**Excessive heat** and **tornado** caused the highest fatalities, whereas **tornado** caused the most injuries in the US from 1995 to 2011.

As for the impact on economy, we have got two sorted lists below by the amount of money cost by damages.  

```{r}
property
crop
```

Below illustrates the total property and crop damage affected by the severe weather events.

```{r}
propertyPlot <- qplot(EVTYPE, data = property, weight = propertyDamage, geom = "bar", binwidth = 1) + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) + scale_y_continuous("Property Damage in US dollars")+ 
    xlab("Severe Weather Type") + ggtitle("Total Property Damage by\n Severe Weather Events in\n the U.S. from 1995 - 2011")

cropPlot<- qplot(EVTYPE, data = crop, weight = cropDamage, geom = "bar", binwidth = 1) + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) + scale_y_continuous("Crop Damage in US dollars") + 
    xlab("Severe Weather Type") + ggtitle("Total Crop Damage by \nSevere Weather Events in\n the U.S. from 1995 - 2011")
grid.arrange(propertyPlot, cropPlot, ncol = 2)
```  

Both **flood** and **hurricane/typhoon** caused the most property damage.
While**drought** and **flood** caused the most crop damage from 1995 to 2011.

### Conclusion  
Based on the findings, **excessive heat** and **tornado** had the greatest impact on the population health, whereas **flood**, **drought**, and **hurricane/typhoon** had the greatest economic impact.


